\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[margin=1in]{geometry}
\usepackage{microtype}
\usepackage{xcolor}
\definecolor{link}{RGB}{14,123,109}
\usepackage[colorlinks=true,urlcolor=link,linkcolor=link]{hyperref}
\pagestyle{empty}
\setlength{\parindent}{0pt}

\begin{document}
\thispagestyle{empty}

\begin{center}
{\large\bfseries Clarifying Without Bias: Chatbots in Experimental Economics}\\[3pt]
{\small Can \c{C}elebi \quad\textperiodcentered\quad \today}
\end{center}
\vspace{0.6em}

A standing threat to inference in experimental economics is that subjects may simply not understand the
task. When choices reflect confusion about the rules or payoffs rather than underlying preferences, the
resulting behavior can mimic the very objects experiments seek to measure: confusion has been shown to
inflate apparent fairness, cooperation, trust, and other-regarding behavior across canonical games, and
to contaminate elicited risk attitudes. Because the same data are used to build theory and to inform
policy, confusion that goes unaddressed distorts both. Experimenters therefore work to clarify
instructions, often relying on research assistants to answer subjects' questions during a session---but
a human assistant is costly, what they tell each subject goes unrecorded and uncontrolled, and turning to
a person can itself introduce bias. This motivates our question: can a large language model take over the assistant's
role, reducing confusion by clarifying the rules without introducing bias by giving advice?

\medskip
Answering it requires a chatbot that clarifies but never steers, and a way to measure how reliably it
stays within that role. We study the question in Oprea's (2024) experiments on choice under risk, where
the interpretation of behavior hinges on separating what subjects \emph{comprehend} from how
\emph{complex} the task is to evaluate. Our chatbot may restate rules and define terms, but is forbidden
to give advice, compute the best option, or reduce the task's complexity. To evaluate it without running
human sessions, we have adversarial ``synthetic subjects''---language models instructed to behave like
difficult participants who seek advice, demand worked examples, or probe for the experimenter's
intent---attack the chatbot over many fifteen-turn conversations, more than 40{,}000 replies in total,
while an independent model scores every reply for advice-giving, factual error, and incoherence. Within
this design we compare alternative chatbot architectures and base models.

\medskip
Two results stand out. First, a \emph{simpler} design is more reliable: a single, well-instructed model
call adheres best, and each additional ``guardrail'' agent we add leaves rule-breaking unchanged or more
common, because the extra steps tend to repeat the subject's forbidden request back to them. Second, the
one change that clearly helps is a \emph{stronger base model}, which cut the bias rate from about 3\% to
under 1\% on the hardest case while also improving accuracy and speed; adding reasoning effort or a
correction pass did not help. Under strict adjudication two of the three error terms nearly vanish:
factual ``errors'' were almost entirely false positives of the first-pass judge, and complexity
violations are rare on any capable model---so the only axis with appreciable signal is \emph{bias}, and
within bias the residual is not outright advice but the bot introducing expected-value or risk language.
Because small per-reply risks accumulate, we report reliability over a whole conversation rather than a
single turn; and because the reported bias rate counts only the conversations a cheap first-pass judge
flagged for review, it is a conservative lower bound. The upshot: for a chatbot meant to clarify without
biasing, the size of the base model matters more than the elaborateness of the surrounding
workflow. Full methods and interactive results:
\href{https://can-celebi.github.io/clarifying-without-bias/}{\texttt{can-celebi.github.io/clarifying-without-bias}}.

\end{document}
